---
title: "Validation: rcisignal vs rcicr infoVal convergence"
date: "Last updated: `r format(Sys.Date(), '%d-%m-%Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Validation: rcisignal vs rcicr infoVal convergence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = TRUE
)
library(rcisignal)

val_cache <- file.path("precomputed", "validation_rcicr")
val_have_cache <- dir.exists(val_cache) &&
  file.exists(file.path(val_cache, "rcicr_per_trait.rds"))
have_rcicr <- requireNamespace("rcicr", quietly = TRUE)
```

## 1. Why this vignette

`rcisignal::infoval()` reimplements the per-producer
informational-value statistic of Brinkman et al. (2019). The
upstream reference implementation is `rcicr::computeInfoVal2IFC()`.
This vignette shows the receipts:

- **2IFC engine equivalence**: the two implementations build the
  same reference distribution under matched conditions, so the
  per-producer z scores they yield are mathematically equivalent.
- **2IFC real-data side-by-side**: on the Oliveira et al. (2019)
  Study 1 data, the two implementations produce ordinally
  consistent per-producer z scores. They diverge in absolute
  magnitude by a predictable, documented amount that comes from
  the two implementations' different default reference-distribution
  choices (per-trial-count vs pool-size).
- **Brief-RC signal-recovery sensitivity**: `rcicr` has no
  Brief-RC `computeInfoVal` path, so no convergence comparison is
  possible. The substitute is a sensitivity check on simulated
  data: planted signal should produce z scores well above 1.96
  while no-signal data should not.

This vignette is the canonical home for the per-producer 2IFC
validation claim in
[`vignette("rcisignal")`](rcisignal.html#validation-status) (§1.2).
For everything else about infoVal (when to use it, how to read
the z, group-mean caveats), read §9 of the main user guide.

## 2. 2IFC engine equivalence

Both `rcisignal::infoval()` and `rcicr::computeInfoVal2IFC()`
build the per-producer reference distribution the same way:
sample random plus-or-minus-one responses against the noise
pool, compute the resulting CI, take its Frobenius norm,
repeat. The two implementations are mathematically equivalent
when `n_trials == n_pool` and `mask = NULL`. rcisignal samples
without replacement (a permutation of the pool) and rcicr
samples with replacement across the fixed-order pool, but the
Frobenius norm is invariant under column permutation, so the
two distributions have the same statistical law.

Demonstrate on a small synthetic noise matrix:

```{r engine-equiv}
set.seed(42L)
n_pix  <- 64L
n_pool <- 50L
iter   <- 1000L

noise <- matrix(stats::rnorm(n_pix * n_pool), n_pix, n_pool)

rcisignal_ref <- rcisignal:::simulate_reference_norms(
  noise_matrix = noise,
  n_trials     = n_pool,
  iter         = iter
)

rcicr_style_ref <- replicate(iter, {
  r <- sample(c(-1, 1), n_pool, replace = TRUE)
  norm((noise %*% r) / n_pool, "f")
})

equiv <- data.frame(
  source = c("rcisignal", "rcicr-style"),
  median = c(stats::median(rcisignal_ref),
             stats::median(rcicr_style_ref)),
  mad    = c(stats::mad(rcisignal_ref),
             stats::mad(rcicr_style_ref))
)
knitr::kable(equiv, digits = 4,
             caption = paste0("Reference-distribution medians and MADs ",
                              "match within Monte Carlo tolerance."))
```

Per-producer z is `(observed_norm - median(ref)) / mad(ref)`,
so matched reference medians and MADs are sufficient for
matched z scores. The corresponding unit test at
`tests/testthat/test-infoval-rcicr-equivalence.R` locks this
in at 2000 iterations and asserts the median gap below 3% and
the MAD gap below 5%.

## 3. 2IFC on Oliveira et al. (2019): per-producer side-by-side

Engine equivalence is the easy half. The harder question is
what the two implementations produce on real data when their
defaults diverge:

- `rcisignal::infoval()` uses a reference distribution matched
  to each producer's actual trial count (300 trials in this
  study).
- `rcicr::computeInfoVal2IFC()` uses a reference distribution
  at the full pool size (4096 in this study), the convention
  from Brinkman et al. (2019).

Because both reference distributions are built from random
sign sequences run through the same mask formula, and the
producer's observed Frobenius norm is the same input on both
sides, the two should agree closely on a 2IFC dataset of this
size. We expect tight per-producer correlations and very
similar (though not identical) absolute z values.

Each per-trait per-participant pair was computed with both
implementations at `iter = 10000` (rcicr's recommended value)
during precompute. The cached output is loaded here:

```{r oliveira-load, eval = val_have_cache}
per_trait <- readRDS(file.path(val_cache, "rcicr_per_trait.rds"))
```

```{r oliveira-summary, eval = val_have_cache}
summary_tbl <- do.call(rbind, lapply(names(per_trait), function(tr) {
  d <- per_trait[[tr]]
  data.frame(
    trait               = tr,
    n_producers         = nrow(d),
    median_z_rcisignal  = stats::median(d$rcisignal_z),
    median_z_rcicr      = stats::median(d$rcicr_z),
    n_ge_1.96_rcisignal = sum(d$rcisignal_z >= 1.96),
    n_ge_1.96_rcicr     = sum(d$rcicr_z >= 1.96),
    pearson_r           = stats::cor(d$rcisignal_z, d$rcicr_z),
    stringsAsFactors    = FALSE
  )
}))
knitr::kable(summary_tbl, digits = 3,
             caption = paste0("Per-trait infoVal side-by-side: ",
                              "median z per implementation, ",
                              "headcount clearing 1.96, ",
                              "per-producer correlation."))
```

```{r oliveira-scatter, eval = val_have_cache, fig.width = 8, fig.height = 6.5, fig.cap = "Per-producer infoVal z scores from rcicr (x-axis) vs rcisignal (y-axis), one panel per Oliveira trait. Dashed lines mark the z = 1.96 threshold; the solid gray line is the identity."}
n_traits <- length(per_trait)
ncol_grid <- 5L
nrow_grid <- ceiling(n_traits / ncol_grid)
op <- par(mfrow = c(nrow_grid, ncol_grid),
          mar   = c(3.2, 3.2, 2.2, 0.8),
          mgp   = c(1.8, 0.5, 0))
on.exit(par(op))
for (tr in names(per_trait)) {
  d <- per_trait[[tr]]
  rng <- range(c(d$rcisignal_z, d$rcicr_z, 0), na.rm = TRUE)
  plot(d$rcicr_z, d$rcisignal_z,
       xlim = rng, ylim = rng,
       xlab = "rcicr z", ylab = "rcisignal z",
       main = tr, pch = 19, cex = 0.8, col = "#1682d4")
  graphics::abline(0, 1, col = "gray60")
  graphics::abline(h = 1.96, v = 1.96, col = "gray80", lty = 3)
}
```

Two patterns to read out of the table and the scatter:

- Pearson r across producers is 1.000 in every trait. The
  rank ordering of producers (and so the headcount clearing
  any fixed threshold) is identical across implementations.
- Median z values differ between the two implementations by
  at most about 0.01 per trait, well within Monte-Carlo noise.
  The headcounts clearing z = 1.96 match exactly.

The takeaway: on this dataset the two implementations are
numerically interchangeable. `infoval()` reimplements the
Brinkman et al. (2019) statistic faithfully; reporting with
either function produces the same conclusions about producer
quality. The only situation where the choice matters is when
prior work in the same line of research has already reported
rcicr-default numbers and you want yours to line up exactly,
in which case call `rcicr::computeInfoVal2IFC()` directly.

```{r oliveira-fallback, eval = !val_have_cache, echo = FALSE, results = "asis"}
cat("> Side-by-side results require the precompute output. ",
    "Run `Rscript data-raw/validation_rcicr/precompute.R` ",
    "with the `rcisignal_OLV_DATA` environment variable set ",
    "to your local copy of the Oliveira (2019) data ",
    "directory. Without the cache this section displays no ",
    "table or figure.\n", sep = "")
```

## 4. Brief-RC: signal-recovery sensitivity

`rcicr` has no Brief-RC `computeInfoVal` path. `rcisignal` ships
a native Brief-RC implementation via [`infoval_report()`] /
[`infoval()`] using a per-trial-count reference distribution. No
convergence comparison against `rcicr` is possible, so this
section runs the next-best demonstration: a sensitivity check on
simulated data.

Plant a strong signal in the eyes region for one simulated
condition, and use a no-signal sim as the negative control.
`infoval()` should clear 1.96 for most producers in the
planted condition and stay near zero in the no-signal
condition.

```{r briefrc-recovery, eval = have_rcicr}
set.seed(11L)
sim_signal <- simulate_briefrc_data(
  n_per_condition = 12L,
  n_trials        = 60L,
  signal_region   = "eyes",
  signal_strength = "strong",
  seed            = 11L
)
sim_noise <- simulate_briefrc_data(
  n_per_condition = 12L,
  n_trials        = 60L,
  signal_strength = "none",
  seed            = 12L
)

cis_signal <- ci_from_responses_briefrc(
  sim_signal$data, noise_matrix = sim_signal$noise_matrix
)
cis_noise <- ci_from_responses_briefrc(
  sim_noise$data, noise_matrix = sim_noise$noise_matrix
)

iv_signal <- infoval(
  cis_signal$signal_matrix, sim_signal$noise_matrix,
  responses = sim_signal$data, iter = 500L,
  seed = 1L, progress = FALSE
)
iv_noise <- infoval(
  cis_noise$signal_matrix, sim_noise$noise_matrix,
  responses = sim_noise$data, iter = 500L,
  seed = 1L, progress = FALSE
)

recovery <- data.frame(
  condition = c("planted (eyes, strong)", "no signal"),
  median_z  = c(stats::median(iv_signal$infoval),
                stats::median(iv_noise$infoval)),
  n_ge_1.96 = c(sum(iv_signal$infoval >= 1.96),
                sum(iv_noise$infoval >= 1.96)),
  n_total   = c(length(iv_signal$infoval),
                length(iv_noise$infoval)),
  stringsAsFactors = FALSE
)
knitr::kable(recovery, digits = 2,
             caption = paste0("Brief-RC infoVal recovers planted signal ",
                              "and returns near-zero z on no-signal data."))
```

```{r briefrc-recovery-plot, eval = have_rcicr, fig.width = 6, fig.height = 4.2, fig.cap = "Per-producer Brief-RC infoVal z. Planted signal (left) concentrates above the 1.96 threshold (dashed); no-signal data (right) centers near zero."}
op <- par(mar = c(4, 4, 2, 1))
on.exit(par(op))
boxplot(list(planted     = iv_signal$infoval,
             "no signal" = iv_noise$infoval),
        ylab = "per-producer z",
        main = "Brief-RC signal-recovery sensitivity",
        col  = c("#1682d4", "gray80"))
graphics::abline(h = 1.96, lty = 2, col = "gray40")
```

```{r briefrc-fallback, eval = !have_rcicr, echo = FALSE, results = "asis"}
cat("> `rcicr` is not installed in this build, so the ",
    "Brief-RC simulator demonstration cannot run ",
    "(`simulate_briefrc_data()` builds its noise pool via ",
    "`rcicr::generateNoisePattern()`). See the source of ",
    "this vignette for the recipe.\n", sep = "")
```

## 5. Conclusion

For 2IFC, the rcisignal and rcicr engines are mathematically
equivalent at the level of the reference distribution and
numerically interchangeable on real data (Oliveira et al.,
2019): per-producer correlation 1.000 in every trait, identical
headcounts clearing z = 1.96, median z values within about 0.01
of each other. For Brief-RC, no rcicr counterpart exists; the
package's native implementation passes a signal-recovery
sanity check on simulated data.

For applied work, `infoval()` is the path forward for both
2IFC and Brief-RC. Call `rcicr::computeInfoVal2IFC()` directly
only when the report needs to sit alongside prior published
work that used the rcicr default.

## References

- Brinkman, L., Goffin, S., van de Schoot, R., van Haren, N. E. M., Dotsch, R., & Aarts, H. (2019). Quantifying the informational value of classification images. *Behavior Research Methods*, *51*(5), 2059-2073. \doi{10.3758/s13428-019-01232-2}
- Dotsch, R. (2016, 2023). *rcicr: Reverse-correlation image-classification toolbox* [R package]. <https://github.com/rdotsch/rcicr>
- Oliveira, M., Garcia-Marques, T., Dotsch, R., & Garcia-Marques, L. (2019). Dominance and competence face to face: Dissociations obtained with a reverse correlation approach. *European Journal of Social Psychology*. \doi{10.1002/ejsp.2569}
- Schmitz, M., Rougier, M., & Yzerbyt, V. (2024). Introducing the brief reverse correlation: An improved tool to assess visual representations. *European Journal of Social Psychology*. \doi{10.1002/ejsp.3100}
